---
title: "Bayesian Data Analysis Project"
output: 
  bookdown::pdf_document2:
    number_sections: false
    toc: FALSE
urlcolor: blue
---

```{r setup, include = TRUE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = TRUE)
library(tidyverse)
theme_set(theme_bw())
require(extraDistr) #need for rdunif
library(dplyr)
```

# Introduction
In our world today, mental health has become more crucial and suicide remains a significant public health concern worldwide, with rates varying across regions due to complex social, economic, and cultural factors. Therefore, understanding the geographic distribution of suicide rates is important for the development of targeted mental health policies and prevention strategies. Although where we live in the world largely affects our mental and physical lifestyles, there has not been many studies done on the geographical characteristics of suicide rates and mental well-being.

In this study, we apply a Bayesian hierarchical model with a conditionally autoregressive (CAR) prior to investigate spatial patterns in suicide rates across countries, stratified by sex groups. We will be using a Poisson likelihood to obtain a prior distribution. After which, inferential analysis will be done using Monte Carlo Markov Chain (MCMC) done in Stan.

This research specifically focuses on two recent years,2019 and 2021, to investigate any observable changes in suicide patterns potentially influenced by global events like the COVID-19 pandemic. Our main research question is: Are there identifiable spatial clusters of high suicide rates, and do these clusters differ by sex? 

This approach allows us to identify high-risk regions, quantify uncertainty, and better understand how neighboring countries may influence each other’s suicide rates, providing valuable insights for data-informed mental health interventions.


# Literature Review


# Dataset and Data Cleaning
Dataset Name: Crude Suicide Rate (Per 100,000 Population)
Source: https://www.who.int/data/gho/data/themes/mental-health/suicide-rates
Description: The raw dataset has notable features like country, age group, sex, and suicide rate (per 100,000 people) that can be extracted.

Location: Country name
Period: Year
Dim1: Sex ("Female", "Both sexes", "Male)
FactValueNumeric: Number of suicide deaths in a year, divided by the population and multiplied by 100 000 (as indicated in the original data source)
FactValueNumericLow: Low estimate 
FactValueNumericHigh: High estimate

Note: The FactValueNumeric data are estimates of the number of suicides. "The estimates are derived from the WHO Global Health Estimates (GHE)" [data source]. However, some countries may not have an accurate way of recording the exact number of deaths, potentially leading to inaccurate estimations. Hence there is a high and low in the death rates.
"For countries without high-quality death registration data, cause of death estimates are calculated using other data, including household surveys with verbal autopsy, sample or sentinel registration systems, special studies" [data source].

```{r}
data_raw = read.csv("suicide_rate_raw.csv", header = TRUE)
head(data_raw)
data = as.data.frame(data_raw |> select(Location, Period, Dim1, FactValueNumeric, FactValueNumericLow, FactValueNumericHigh))
head(data)
unique(data$Period)
```
filter data to only include 2019 and 2021
```{r}
suicide_2019 <- filter(data, Period == 2019)
suicide_2021 <- filter(data, Period == 2021)
```


# Data Analysis
As we have obtained the cleaned data for suicide rates in 2019 and 2021, we can now declare a prior model from information obtained historically.




