---
title: "Bayesian Data Analysis Project"
output: 
  bookdown::pdf_document2:
    number_sections: false
    toc: FALSE
urlcolor: blue
---

```{r setup, include = TRUE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = TRUE)
library(tidyverse)
theme_set(theme_bw())
require(extraDistr) #need for rdunif
library(dplyr)
```

# Introduction
In our world today, mental health has become more crucial and suicide remains a significant public health concern worldwide, with rates varying across regions due to complex social, economic, and cultural factors. Therefore, understanding the geographic distribution of suicide rates is important for the development of targeted mental health policies and prevention strategies. Although where we live in the world largely affects our mental and physical lifestyles, there has not been many studies done on the geographical characteristics of suicide rates and mental well-being.

In this study, we apply a Bayesian hierarchical model with a conditionally autoregressive (CAR) prior to investigate spatial patterns in suicide rates across countries, stratified by sex groups. We will be using a Poisson likelihood to obtain a prior distribution. After which, inferential analysis will be done using Monte Carlo Markov Chain (MCMC) done in Stan.

This approach allows us to identify high-risk regions, quantify uncertainty, and better understand how neighboring countries may influence each other’s suicide rates, providing valuable insights for data-informed mental health interventions.


# Literature Review


# Dataset and Data Cleaning
Dataset Name: Crude Suicide Rate (Per 100,000 Population)
Source: https://www.who.int/data/gho/data/themes/mental-health/suicide-rates
Description: The raw dataset has notable features like country, age group, sex, and suicide rate (per 100,000 people) that can be extracted.

Location: Country name
Period: Year 2019
Dim1: Sex ("Female", "Both sexes", "Male)
Dim2: Age group (15-29 years, 30-49 years)

```{r}
data_raw = read.csv("old_young.csv", header = TRUE)
head(data_raw)
data = as.data.frame(data_raw |> select(Location, Period, Dim1, Dim2))
head(data)
unique(data$Dim2)
```
Separate dataset into two age groups: "15-29 years" and "30-49 years"
```{r}
suicide_rate_1529 <- data |> filter(Dim2 == "15-29  years") #THERES AN EXTRA SPACE

suicide_rate_3049 <- data |> filter(Dim2 == "30-49 years")

suicide_rate_1529 #CAN BE FILTERED INTO EACH SEX TOO
suicide_rate_3049

```

# Data Analysis





